Landau quantization of nearly degenerate bands

### 2018-09-29 paper.tex
Updates from ShareLaTeX

In the absence of magnetic breakdown, semiclassical quantization rules are well-known for Landau levels of  both nondegenerate [Onsager, Lifshitz] and degenerate [Roth, Mikitik, Topoferm, 100page] energy bands. Our work generalizes these rules to nearly-degenerate bands, whose orbits are split in $\bk$-space with a differential area  $\delta S$ that is much less than the average area $S$ of the orbits; we shall refer to $\delta S/S{\ll}1$ as the \textit{quasidegeneracy} condition. Quasidegenerate orbits commonly originate from spin-orbit-splitting of energy bands in noncentrosymmetric solids. We provide an illustrative model of  a 2DEG with Rashba spin-orbit coupling:
\begin{equation}
H_R=\frac{{\hbar^2} k^2}{2m}+\alpha  (k_{x}\sigma_{y}-k_{y}\sigma_{x}).\label{eq:Rashba-Hamiltonian}
\end{equation}
At energies much greater than $m\alpha^2$, the differential area of the two spin-orbit-split orbits [see Fig. \ref{fig:orbits}(a)] is much smaller than their average area:\footnote{The expressions for $\delta S$ and $S$ in \q{deltaSvsS} are asymptotically valid for $E{\gg}m\alpha^2$.}
\e{\delta S = 2\pi k_E \f{2\alpha m}{\hbar^2} \ll S =\pi k_E^2; \;\; k_{E}:=\f{\sqrt{2m E}}{\hbar} .\la{deltaSvsS}}
$S$ is equivalently the area of the circular, constant-energy contour in the absence of spin-orbit coupling ($\alpha{=}0$).

@@

Let us heuristically argue for the form of such a rule; a more pedagogical and rigorous approach is elaborated in App. \ref{app:quantizationruleproof}. We consider independent electrons described (at zero field) by a translation-invariant Hamiltonian: $\hat{H}_0{+}\delta \hat{H}$. This decomposition is such that energy bands of $\hat{H}_0$ are $D$-fold degenerate at every crystal wavevector $\bk$, while $\delta \hat{H}$ is a weak perturbation that lifts this degeneracy at generic $\bk$. For conceptual simplicity, we focus on $D{=}2$ spin degeneracy, and postpone the general formulation for $D{\geq}2$ to App. \ref{app:quantizationruleproof}.  $\hat{H}_0$ is commonly exemplified by  the Schr\"odinger Hamiltonian, and $\delta \hat{H}$ by a non-centrosymmetric spin-orbit interaction; alternatively, $\hat{H}_0$ may be a centrosymmetric Pauli Hamiltonian (with spin-orbit coupling), while $\delta H$ describes a lattice deformation that breaks the spatial-inversion symmetry.

% The quantum theory of a Bloch electron in a magnetic field is based on the noncommutivity of the kinematic quasimomentum operators:[onsager, lifshitz]

@@

To the first order, $\delta \hat{H}$ and the Zeeman interaction induce a subleading phase offset ($\lambda_a$) which generically differs for the two orthogonal spinor states (labelled $a{=}1,2$):
\begin{equation}
l^2S(E)+\lambda_a(E,B)+\gamma=2\pi n,\as  n\in \mathbb{Z},\;a=1,2. \label{eq:rule}
\end{equation}
This implies that the spin-degeneracy of Landau levels is split   by approximately  $\var_c|\lambda_1{-}\lambda_2|/2\pi$. The phase offsets $\lambda_a$ can be viewed as Floquet quasienergies of a time-periodic, perturbative Hamiltonian ($\calh$) that governs the dynamics of the spinor wavefunction.  $\lambda_a$ is obtained  from  eigenvalues $\{e^{i\lambda_a}\}_{a=1}^2$ of a time-evolution propagator -- the time-ordered exponential 
\begin{equation}
\A{:}{=}\overline{\exp}\left[-\frac{i}{\hbar}\int_0^{T_c} \calh(\bk(t)) dt\right],
\label{eq:prop}
\end{equation} 
with $\bk(t){\in}\frako_0$ a periodic function determined by the semiclassical equation of motion.   $\calh$ has the two-by-two matrix form:
\begin{equation}
  \calh=\delta \var +B\bigg(M_z -g_s\mu_{B}\f{s_z}{\hbar} + e\epsilon_{\alpha\beta}\mathfrak{X}_{\beta}v_{\alpha}\bigg):=\delta \var+B\calm,\label{eq:H1}
\end{equation}
$\delta \var$ is the projection of $\delta H$ to the quasidegenerate band associated to $\var(\bk)$, e.g., $\delta \var{=}\alpha(k_x\sigma_y{-}k_y\sigma_x)$ in the Rashba model. In the basis of energy-split bands (which we label by ${+}$ and $-$ for bands corresponding to the outer and inner orbit respectively), $\delta \var$ is a diagonal matrix whose diagonal entries ($\delta \var_+,\delta \var_-$, corresponding to outer and inner orbit respectively; $\delta \var_+<\delta \var_-$) are related to the hybridization parameter:\footnote{The neglected term on the right-hand side of  \q{spinsplitwkb} is smaller (than the kept term) by a factor $\delta v_{\pm}/v$, which is the relative change in band speed $v=|\boldsymbol{v}|=\hbar^{-1}|\partial \var/\partial \bk|$ induced by the spin-orbit coupling. This factor is assumed to be small.  For example in the Rashba model, $\delta v_{\pm}/v=m\alpha/\hbar k_E{\sim}\delta S/S{\ll}1$.}   
\e{-\int_0^{T_c} (\delta \var_+{-}\delta \var_-) \frac{dt}{\hbar} \approx l^2\delta S. \la{spinsplitwkb}}
%In the  Rashba model,   $\delta \var_{\pm}(E){=}{\pm} (\alpha/\hbar)\sqrt{2mE}$ is independent of $\bk$ due to a continuous rotational symmetry of $H_R$. \\

@@

%With intermediate hybridization , including the regime where Landau-Zener dynamics is non-applicable and breakdown is nonlocal.

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RZ.png}
\caption{(a) Trefoille-shaped cyclotron orbit for the Rashba-Zeeman model at the energy of the Dirac point. The orbit is drawn over the  spin-splitting field ($d_x,d_y$) of the Hamiltonian $H_{RZ}{=}d_0(\bk)I{+}d_x(\bk)\sigma_x{+}d_y(\bk)\sigma_y$, with $d_j$ obtainable from \q{eq:RZ-Hamiltonian}. The wavefunction along the inner (resp.\ outer) component of the trefoille orbit has its pseudospin aligned parallel (resp.\ anti-parallel) to the spin-orbit field. (b) Oscillation of $\delta E_n$ [cf. Eq. (\ref{RZfirstorderE})] for even integer $n$ with respect to magnetic field calculated by Eq. (\ref{RZfirstorderE}) (black dashed line) and numerical diagonalization (red solid line). $\delta E$ is extracted at 0.45eV and the II-Dirac point is located at 0.5eV. Parameters for (b): $m$: 0.076$m_e$, $\alpha$: 1.5($10^{5}$cm/s$\cdot\hbar$), $Z_\parallel$: 1meV. Out of plane Zeeman coupling is ignored here. (c) A sketch of cyclotron orbits of Rashba Hamiltonian with in-plane magnetic field. (d) For various configurations ($\alpha$, $Z_\parallel$) of $H_{RZ}$, we plot in left panel $\text{error}:=|\epsilon_\text{rule}-\epsilon_\text{exact}|/\var_c$ for a single Landau level that is closest to $\epsilon_0$ in Eq. (\ref{whereisdiracpoint}), where $\epsilon_\text{rule}$ is determined by Eq. (\ref{eq:rule}) and $\epsilon_{\text{exact}}$ is obtained by numerical diagonalization of the Peierls-Onsager Hamiltonian $H_{RZ}(\bK)$. Corresponding $l^2 \delta S^2/S$ are plotted in right panel. We only plotted in region where $0.4<\epsilon_0<1.0$(eV).\cite{RZfignote}, Units: $\alpha$($1.5\times 10^{5}$cm/s$\cdot\hbar$). Parameters for (d): $m=0.076m_e$, $l=100$\AA. \label{fig:RZ}}.
\end{figure}

@@

As alluded to in \s{sec:qtznrules}, the adiabatic limit of two independent orbits does not exist if the two spin-orbit-split orbits exactly intersect at a II-Dirac point. To investigate this failure of adiabaticity, let us study the Landau levels at energies close to the II-Dirac point ($\epsilon_0$). We focus first on the regime where, on average (over $\frako_0$), the energy splitting $|\delta \var_+{-}\delta \var_-|$ [cf.\ \q{eq:RZ-Hamiltonian}] is much greater than $B_\perp|\calm_{\sma{+-}}|$. Equivalently, we are in the regime where $\alpha k_{\sma{E}}{\gg}\var_c$.\footnote{The orbit-averaged energy splitting is of order $\alpha k_{\sma{E}}$ for energies near the Dirac point. $B_\perp|\calm_{\sma{+-}}|$ is simply the Berry contribution to the Zeeman interaction, and has the form of the right-most term in \q{rashbaeffham}; this term is of order $\var_c$.} This implies that the split orbits do not appreciably hybridize -- except in the vicinity of the II-Dirac point where orbits (nearly) touch.  Near this touching point (where $t{=}0$), the dynamics of a spinor electron is described by linearizing $\H$ [cf.\ \q{calHRZ}] with respect to $t$:
\e{
\calh = -\alpha k_{E}\omega^{\sma{\perp}}_c t\,\sigma_x+(Z_\parallel-\alpha k_{E})\sigma_y-Z_\perp\sigma_z,\la{effhamIIDirac}
}
The eigenbasis of $\sigma_x$ corresponds to two diabatic levels with characteristic slope $v_d{:}{=}\alpha k_E\omega^{\sma{\perp}}_c$; near $t{=}0$, the diabatic levels anticross  due to the hybridization terms proportional to $\sigma_{y,z}$. Generally, the probability of tunneling across the hybridization-induced barrier is  $\rho^2{=}\exp(-2\pi\barmu)$, with $\barmu{=}{E_g}^2/{2v_d \hbar}$ and $E_g$ the energy gap at the center of the anticrossing;[Landau and lifshitz, Wittig] in this context,
\e{\barmu=\frac{Z_\perp^2+(Z_\parallel-\alpha k_{E})^2}{2\alpha k_{E}\var^{\sma{\perp}}_c}, \la{mu1}}
with $2\alpha k_E$ the spin-orbit splitting at energy $E$ [cf.\ \fig{fig:orbits}(a)]. In the absence of the out-of-plane Zeeman coupling ($Z_\perp{=}0$), $\bar{\mu}$ would vanish at the energy of the II-Dirac point -- leading to unit-probability Klein tunneling, as was first proposed by [Obrien] in a different model of the II-Dirac point. However, with a proper account of $Z_\perp$, we find instead that $\bar{\mu}$ has a nonvanishing minimum: $\bar{\mu}_{\sma{\text{min}}}{:}{=} ( m/m_0)(Z_\perp/4\alpha k_{E})$, which is a product of the dimensionless effective mass and the ratio of the out-of-plane Zeeman splitting over the spin-orbit splitting. This minimum is practically negligible for  semiconductor heterostructures with $m{\ll}m_0$, but may be relevant in heavy-fermion systems.

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{blochsphere.png}
\caption{Solid-angle representation of  the single-band Berry phase $\phi_{\pm}^B$ for inner (subscript $-$, upper panel) and outer ($+$, lower panel) orbit, at various energies close to the Dirac point $\epsilon_0$ [cf.\ Eq. (\ref{whereisdiracpoint})]. The decomposition of the effective Hamiltonian [cf.\ Eq. (\ref{calHRZ})] into sums of Pauli matrices ($\calh{=}\sum_jd_j\sigma_j$) defines a three-vector $\bd{=}({-}\alpha k_y$,$Z_\parallel{+}\alpha k_x,{-}Z_\perp$).
 $\phi_{\pm}^B$  equals half the solid angle subtended by   ${\pm} \bd(\bk)$ as $\bk$ is varied along the outer/inner orbit.[Berry]   These angles are indicated in the figure by blue outlines.\label{fig:blochsphere}}
\end{figure}

@@

For any nonzero field, $0{<}\tau{<}1$ implies that the two incommensurate harmonics $(\Omega_+{\pm}\Omega_-)$ in \q{coscos} compete to produce a quasirandom Landau-level spectrum [kaganovslutskin]. That is to say, the spectrum is disordered on the scale of nearest-neighbor level spacings but has longer-ranged correlations. To appreciate this, let us analyze the Landau levels (near the Dirac point) perturbatively in the small parameter $\tau$. To zeroth order, \q{coscos} reduces to
\e{2l^2S(E_n^0)+\pi = 2\pi n \rightarrow \{E_n^0(B)\}_{n\in\Z}.\label{trefoillerule}}
To derive the above equation, we have utilized $\phi^B_-{+}\phi^B_+{=}0$, which is valid for any two-band model in the quasidegenerate regime\footnote{For two band models, Berry phases of the two bands sum up to 0 on arbitrary path}; in \fig{fig:blochsphere} this is reflected in the sum of solid angles (at fixed energy) equalling $4\pi$. \q{trefoillerule} may be interpreted as a quantization rule for a trefoille-shaped orbit with net area $2S{=}4\pi m E{/\hbar^2}$; the corresponding Landau levels are periodic in $E{\rightarrow}E{+}\pi/l^2(\partial S/\partial E)$ and in $l^2{\rightarrow}l^2{+}\pi/S$. Both periods are approximately half that of either disconnected  orbit, in the assumed quasidegenerate regime $\delta S{\ll}S$. The $\pi$ correction in \q{trefoillerule} is the Berry phase associated to a $2\pi$ rotation of the wavefunction spin over a single trefoille orbit, as illustrated in Fig. \ref{fig:RZ}(a). The Maslov phase, which is generally equal to $\pi$ times the rotation number of an orbit, is trivially $2\pi$ for the trefoille orbit.

%To appreciate this, consider the regime $\rho^2{\lesssim}1$, where a single harmonic (corresponding to the trefoille orbit) is dominant. Viewing $\tau{=}\sqrt{1{-}\rho^2}$ as a small parameter, the zeroth-order Landau levels $E_n^0(B)$ [as obtained from \q{trefoillerule}] has been shown to be periodic in $E$ and $l^2$ (on short scales associated to the large-action trefoille orbit).

To recapitulate, $E_n^0(B)$ [cf.\ \q{trefoillerule}] exhibits a periodicity on short scales associated to the large area of the trefoille orbit. This short-scale periodicity is lost due to  finite-$\tau$ corrections, yet a longer-range correlation (associated to the smaller differential action $\delta S$ between spin-split orbits) persists, as shown by the first-order-in-$\tau$ correction:\footnote{The general form of this perturbative calculation can be found in Sec. IX-E of [100page].} 
\e{&\delta E_n(B) = \f{(-1)^{n+1}}{2\pi} \tau \var_c \cos\bigg(\f{l^2\delta S}{{2}}+\f{\phi_+^B-\phi_-^B}{2}-\bar{\varphi}\bigg) \lin
    &\approx \f{(-1)^{n+1}}{4\sqrt{2\pi}}  \var_c \f{(E-\epsilon_0)l}{\alpha(\epsilon_0/Z_{\parallel})^{3/2}}\cos\bigg(\f{l^2\delta S}{{2}}-\bar{\varphi}\bigg),\label{RZfirstorderE}}
with the right-hand side (of each line) evaluated at $E{=}E_n^0(B)$. The second line is an approximation for $Z_{\perp}{\rightarrow}0$, and has greater utility where $m{\ll}m_0$. \red{Eq. (\ref{RZfirstorderE}) are compared with numerical diagonalization Fig. \ref{fig:RZ}(b) with even integers $n$.}

%( $|\delta \var_+{-}\delta \var_-|{\lesssim}B_\perp|\calm_{\sma{+-}}|$ on average)

@@

%For the special case of $\alpha{=}0$, the spectra of $H_{RZ}(\boldsymbol{K})$ and $\mathcal{A}[\mathfrak{o}]$ are both exactly soluble, hence it is possible to verify that the exact and semiclassical Landau levels match perfectly for every level index $n$ without any approximation; the latter is obtained from our rule [Eq.\ (\ref{eq:rule})], with the eigenphases of $\A$ equal to 
%\e{\lambda_{\pm}{=}{\pm} 2\pi ml^2\sqrt{\f{Z_\parallel^2}{\hbar^4}{+}\f{g_\text{eff}^2}{16m^2l^4}}.}

%Together with Eq. (\ref{eq:rule}), $\lambda_\pm$ here generate exact the same Landau level spectrum as $H_{RZ}(\boldsymbol{K})$.

%The above quantization rule is not applicable in the nonlocal-breakdown regime ($\Delta/\epsilon_c\sim1$). The appropriate rule is summarized in \qq{eq:rule}{eq:H1}, with $\delta \var$ encoding the change in band energies due to the spin-orbit coupling ($\sim \alpha$) and the in-plane Zeeman coupling ($\sim Z$); $\alpha,Z$ and $B$ are considered  small parameters.

% Thus quasidegenerate quantization rule produces exact solution without Rashba spin-orbit coupling. After Rashba spin-orbit coupling is switched on, no analytic solution of $H_{RZ}(\boldsymbol{K})$ is known to our knowledge. Again, exact Landau levels are solved numerically and are compared with quantization rules (see Fig. \ref{fig:errors} (d-f)). Although $\alpha\ne 0$ in Fig. \ref{fig:errors} (f), accuracy of quantization rules is still largely independent of the choice of $Z$.

@@

We have argued in \s{sec:qtznrules} that the subleading corrections ($\lambda_{1,2}$) to the quantization rule [cf.\ \q{eq:rule}] generically lift the spin-degeneracy of Landau levels -- the level splitting is approximately $\var_c(\lambda_1{-}\lambda_2)/2\pi$. Conversely, Landau levels are approximately spin-degenerate if $\lambda_1{=}\lambda_2$ (mod $2\pi$) -- this degeneracy condition for the eigenvalues of $\A$ may require fine-tuning of the Peierls-Onsager Hamiltonian.

@@

In this and the subsequent section [\s{sec:rotsymmbreakdown}], we will demonstrate how rotational symmetry reduces codimension in the Landau-level problem. Before delving into this, it is instructive to consider the conceptually-simplest analog in differential geometry: if we parametrize the 2D plane by Cartesian coordinates $(x,y){\in}\R^2$, it is apparent that two parameters must be tuned to arrive at the origin. If however the origin is the center of a rotation symmetry, we may instead utilize polar coordinates and tune a single parameter (the radial coordinate) to arrive at $x{=}y{=}0$. We see that rotational symmetry reduces the codimension of the origin from two to one. Here, and in the  Landau-level problem, we define codimension with respect to a \textit{symmetrically-chosen coordinates} (if symmetry exists); such symmetric coordinates need not be well-defined at isolated high-symmetry points, just as the polar coordinates are not well-defined at the rotational center.

>>>>

In the absence of magnetic breakdown, semiclassical quantization rules are well-known for Landau levels of  both nondegenerate [Onsager, Lifshitz] and degenerate [Roth, Mikitik, Topoferm, 100page] energy bands. Our work generalizes these rules to nearly-degenerate bands, whose orbits are split in $\bk$-space with a differential area  $\delta S$ that is much less than the average area $S$ of the orbits; we shall refer to $\delta S/S{\ll}1$ as the \textit{quasidegeneracy} condition. Quasidegenerate orbits commonly originate from spin-orbit-splitting of energy bands in noncentrosymmetric solids. We provide an illustrative model of a two dimensional electron gas (2DEG) with Rashba spin-orbit coupling:
\begin{equation}
H_R=\frac{{\hbar^2} k^2}{2m}+\alpha  (k_{x}\sigma_{y}-k_{y}\sigma_{x}).\label{eq:Rashba-Hamiltonian}
\end{equation}
At energies much greater than $m\alpha^2$, the differential area of the two spin-orbit-split orbits [see Fig. \ref{fig:orbits}(a)] is much smaller than their average area:\footnote{The expressions for $\delta S$ and $S$ in \q{deltaSvsS} are asymptotically valid for $E{\gg}m\alpha^2$.}
\e{\delta S = 2\pi k_E \f{2\alpha m}{\hbar^2} \ll S =\pi k_E^2; \;\; k_{E}:=\f{\sqrt{2m E}}{\hbar} .\la{deltaSvsS}}
$S$ is equivalently the area of the circular, constant-energy contour in the absence of spin-orbit coupling ($\alpha{=}0$).

@@

Let us heuristically argue for the form of such a rule; a more pedagogical and rigorous approach is elaborated in App. \ref{app:quantizationruleproof}. We consider independent electrons described (at zero field) by a translation-invariant Hamiltonian: $\hat{H}_0{+}\delta \hat{H}$. This decomposition is such that energy bands of $\hat{H}_0$ are $D$-fold degenerate at every crystal wavevector $\bk$, while $\delta \hat{H}$ is a weak perturbation that lifts this degeneracy at generic $\bk$. For conceptual simplicity, we focus on $D{=}2$ spin degeneracy, and postpone the general formulation for $D{\geq}2$ to App. \ref{app:quantizationruleproof}.  $\hat{H}_0$ is commonly exemplified by  the Schr\"odinger Hamiltonian, and $\delta \hat{H}$ by a non-centrosymmetric spin-orbit interaction; alternatively, $\hat{H}_0$ may be a centrosymmetric Pauli Hamiltonian (with spin-orbit coupling), while $\delta \hat{H}$ describes a lattice deformation that breaks the spatial-inversion symmetry.

@@

To the first order, $\delta \hat{H}$ and the Zeeman interaction induce a subleading phase offset ($\lambda_a$) which generically differs for the two orthogonal spinor states (labelled $a{=}1,2$):
\begin{equation}
l^2S(E)+\lambda_a(E,B)+\gamma=2\pi n,\as  n\in \mathbb{Z},\;a=1,2. \label{eq:rule}
\end{equation}
This implies that the spin-degeneracy of Landau levels is split   by approximately  $\var_c|\lambda_1{-}\lambda_2|/2\pi$. The phase offsets $\lambda_a$ can be viewed as Floquet quasienergies of a time-periodic, perturbative Hamiltonian ($\calh$) that governs the dynamics of the spinor wavefunction.  $\lambda_a$ is obtained  from  eigenvalues $\{e^{i\lambda_a}\}_{a=1}^2$ of a time-evolution propagator -- the time-ordered exponential 
\begin{equation}
\A{:}{=}\overline{\exp}\left[-\frac{i}{\hbar}\int_0^{T_c} \calh(\bk(t)) dt\right],
\label{eq:prop}
\end{equation} 
with $\bk(t){\in}\frako_0$ a periodic function determined by the semiclassical equation of motion.   $\calh$ has the two-by-two matrix form:
\begin{equation}
  \calh=\delta \var +B\bigg(M_z -g_s\mu_{B}\f{s_z}{\hbar} + e\epsilon_{\alpha\beta}\mathfrak{X}_{\beta}v_{\alpha}\bigg):=\delta \var+B\calm,\label{eq:H1}
\end{equation}
$\delta \var$ is the projection of $\delta \hat{H}$ to the degenerate band subspace of $\hat{H}_0$. In the basis of energy-split bands (which we label by ${+}$ and $-$ for bands corresponding to the outer and inner orbit respectively), $\delta \var$ is a diagonal matrix whose diagonal entries ($\delta \var_+,\delta \var_-$, corresponding to outer and inner orbit respectively; $\delta \var_+<\delta \var_-$) are related to the hybridization parameter:\footnote{The neglected term on the right-hand side of  \q{spinsplitwkb} is smaller (than the kept term) by a factor $\delta v_{\pm}/v$, which is the relative change in band speed $v=|\boldsymbol{v}|=\hbar^{-1}|\partial \var/\partial \bk|$ induced by the spin-orbit coupling. This factor is assumed to be small.  For example in the Rashba model, $\delta v_{\pm}/v=m\alpha/\hbar k_E{\sim}\delta S/S{\ll}1$.}   
\e{-\int_0^{T_c} (\delta \var_+{-}\delta \var_-) \frac{dt}{\hbar} \approx l^2\delta S. \la{spinsplitwkb}}
%In the  Rashba model,   $\delta \var_{\pm}(E){=}{\pm} (\alpha/\hbar)\sqrt{2mE}$ is independent of $\bk$ due to a continuous rotational symmetry of $H_R$. \\

@@

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RZ.png}
\caption{(a) Trefoille-shaped cyclotron orbit for the Rashba-Zeeman model at the energy of the Dirac point. The orbit is drawn over the  spin-splitting field ($d_x,d_y$) of the Hamiltonian $H_{RZ}{=}d_0(\bk)I{+}d_x(\bk)\sigma_x{+}d_y(\bk)\sigma_y$, with $d_j$ obtainable from \q{eq:RZ-Hamiltonian}. The wavefunction along the inner (resp.\ outer) component of the trefoille orbit has its pseudospin aligned parallel (resp.\ anti-parallel) to the spin-orbit field. (b) Upper panel: a sketch of Landau level dispersion near II-Dirac point of Rashba model with in plane Zeeman field in the weak hybridization limit. The dashed line representing first order solution [cf. Eq. (\ref{RZfirstorderE})] oscillates around the solid line [zeroth order solution, cf. Eq. ({trefoillerule})]. Due to the smallness of $\delta S$, the oscillation should not be observable in the scale of several Landau levels. Therefore, We exaggerate the oscillation frequency in the sketch. Lower panel: oscillation of $\delta E_n$ [cf. Eq. (\ref{RZfirstorderE})] for even integer $n$ with respect to magnetic field calculated by Eq. (\ref{RZfirstorderE}) (black dashed line) and numerical diagonalization (red solid line). $\delta E$ is extracted at 0.45eV and the II-Dirac point is located at 0.5eV. Parameters for lower panel of (b): $m$: 0.076$m_e$, $\alpha$: 1.5($10^{5}$cm/s$\cdot\hbar$), $Z_\parallel$: 1meV. Out of plane Zeeman coupling is ignored here. (c) A sketch of cyclotron orbits of Rashba Hamiltonian with in-plane magnetic field. Inset: illustration of $S_{\sma{\square}}$ by the purple rectangle. (d) For various configurations ($\alpha$, $Z_\parallel$) of $H_{RZ}$, we plot in left panel $\text{error}:=|\epsilon_\text{rule}-\epsilon_\text{exact}|/\var_c$ for a single Landau level that is closest to $\epsilon_0$ in Eq. (\ref{whereisdiracpoint}), where $\epsilon_\text{rule}$ is determined by Eq. (\ref{eq:rule}) and $\epsilon_{\text{exact}}$ is obtained by numerical diagonalization of the Peierls-Onsager Hamiltonian $H_{RZ}(\bK)$. Corresponding $l^2 \delta S^2/S$ are plotted in right panel. We only plotted in region where $0.4<\epsilon_0<1.0$(eV).\cite{RZfignote}, Units: $\alpha$($1.5\times 10^{5}$cm/s$\cdot\hbar$). Parameters for (d): $m=0.076m_e$, $l=100$\AA. \label{fig:RZ}}.
\end{figure}

@@

As alluded to in \s{sec:qtznrules}, the adiabatic limit of two independent orbits does not exist if the two spin-orbit-split orbits exactly intersect at a II-Dirac point. To investigate this failure of adiabaticity, let us study the Landau levels at energies close to the II-Dirac point ($\epsilon_0$). We focus first on the regime where, on average (over $\frako_0$), the energy splitting $|\delta \var_+{-}\delta \var_-|$ [cf.\ \q{eq:RZ-Hamiltonian}] is much greater than $B_\perp|\calm_{\sma{+-}}|$. Equivalently, we are in the regime where $\alpha k_{\sma{E}}{\gg}\var_c^\perp$.\footnote{The orbit-averaged energy splitting is of order $\alpha k_{\sma{E}}$ for energies near the Dirac point. $B_\perp|\calm_{\sma{+-}}|$ is simply the Berry contribution to the Zeeman interaction, and has the form of the right-most term in \q{rashbaeffham}; this term is of order $\var_c$.} This implies that the split orbits do not appreciably hybridize -- except in the vicinity of the II-Dirac point where orbits (nearly) touch.  Near this touching point (where $t{=}0$), the dynamics of a spinor electron is described by linearizing $\H$ [cf.\ \q{calHRZ}] with respect to $t$:
\e{
\calh = -\alpha k_{E}\omega^{\sma{\perp}}_c t\,\sigma_x+(Z_\parallel-\alpha k_{E})\sigma_y-Z_\perp\sigma_z,\la{effhamIIDirac}
}
The eigenbasis of $\sigma_x$ corresponds to two diabatic levels with characteristic slope $v_d{:}{=}\alpha k_E\omega^{\sma{\perp}}_c$; near $t{=}0$, the diabatic levels anticross  due to the hybridization terms proportional to $\sigma_{y,z}$. Generally, the probability of tunneling across the hybridization-induced barrier is  $\rho^2{=}\exp(-2\pi\barmu)$, with $\barmu{=}{E_g}^2/{2v_d \hbar}$ and $E_g$ the energy gap at the center of the anticrossing;[Landau and lifshitz, Wittig] in this context,
\e{\barmu=\frac{Z_\perp^2+(Z_\parallel-\alpha k_{E})^2}{2\alpha k_{E}\var^{\sma{\perp}}_c}, \la{mu1}}
with $2\alpha k_E$ the spin-orbit splitting at energy $E$ [cf.\ \fig{fig:orbits}(a)]. $\bar{\mu}$ has an alternative interpretation as the area of rectangle inscribed between the two hyperbolic arms $S_{\sma{\square}}$, as illustrated in the inset of Fig. \ref{fig:RZ}(c). In the absence of the out-of-plane Zeeman coupling ($Z_\perp{=}0$), $\bar{\mu}$ would vanish at the energy of the II-Dirac point -- leading to unit-probability Klein tunneling, as was first proposed by [Obrien] in a different model of the II-Dirac point. However, with a proper account of $Z_\perp$, we find instead that $\bar{\mu}$ has a nonvanishing minimum: $\bar{\mu}_{\sma{\text{min}}}{:}{=} ( m/m_0)(Z_\perp/4\alpha k_{E})$, which is a product of the dimensionless effective mass and the ratio of the out-of-plane Zeeman splitting over the spin-orbit splitting. This minimum is practically negligible for  semiconductor heterostructures with $m{\ll}m_0$, but may be relevant in heavy-fermion systems.

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{blochsphere.png}
\caption{Solid-angle representation of  the single-band Berry phase $\phi_{\pm}^B$ for inner (subscript $-$, upper panel) and outer ($+$, lower panel) orbit, at various energies close to the Dirac point $\epsilon_0$ [cf.\ Eq. (\ref{whereisdiracpoint})]. The decomposition of the effective Hamiltonian [cf.\ Eq. (\ref{calHRZ})] into sums of Pauli matrices ($\calh{=}\sum_jd_j\sigma_j$) defines a three-vector $\bd{=}({-}\alpha k_y$,$Z_\parallel{+}\alpha k_x,{-}Z_\perp$).
 $\phi_{\pm}^B$  equals half the solid angle subtended by   ${\mp} \bd(\bk)$ as $\bk$ is varied along the outer/inner orbit.[Berry]   These angles are indicated in the figure by blue outlines.\label{fig:blochsphere}}
\end{figure}

@@

For any nonzero field, $0{<}\tau{<}1$ implies that the two incommensurate harmonics $(\Omega_+{\pm}\Omega_-)$ in \q{coscos} compete to produce a quasirandom Landau-level spectrum [kaganovslutskin]. That is to say, the spectrum is disordered on the scale of nearest-neighbor level spacings but has longer-ranged correlations. To appreciate this, let us analyze the Landau levels (near the Dirac point) perturbatively in the small parameter $\tau$. To zeroth order, \q{coscos} reduces to
\e{2l^2S(E_n^0)+\pi = 2\pi n \rightarrow \{E_n^0(B)\}_{n\in\Z}.\label{trefoillerule}}
To derive the above equation, we have utilized $\phi^B_-{+}\phi^B_+{=}0$, which is valid for any two-band model in the quasidegenerate regime\footnote{For two band models, Berry phases of the two bands sum up to 0 on arbitrary path}; in \fig{fig:blochsphere} this is reflected in the sum of solid angles (at fixed energy) equalling $4\pi$. \q{trefoillerule} may be interpreted as a quantization rule for a trefoille-shaped orbit with net area $2S{=}4\pi m E{/\hbar^2}$; the corresponding Landau levels (sketched by solid lines in the upper panel of Fig. \ref{fig:RZ}(b)) are periodic in $E{\rightarrow}E{+}\pi/l^2(\partial S/\partial E)$ and in $l^2{\rightarrow}l^2{+}\pi/S$. Both periods are approximately half that of either disconnected  orbit, in the assumed quasidegenerate regime $\delta S{\ll}S$. The $\pi$ correction in \q{trefoillerule} is the Berry phase associated to a $2\pi$ rotation of the wavefunction spin over a single trefoille orbit, as illustrated in Fig. \ref{fig:RZ}(a). The Maslov phase, which is generally equal to $\pi$ times the rotation number of an orbit, is trivially $2\pi$ for the trefoille orbit.

To recapitulate, $E_n^0(B)$ [cf.\ \q{trefoillerule}] exhibits a periodicity on short scales associated to the large area of the trefoille orbit. This short-scale periodicity is lost due to  finite-$\tau$ corrections, yet a longer-range correlation (associated to the smaller differential action $\delta S$ between spin-split orbits) persists, as shown by the first-order-in-$\tau$ correction:\footnote{The general form of this perturbative calculation can be found in Sec. IX-E of [100page].} 
\e{&\delta E_n(B) = \f{(-1)^n}{2\pi} \tau \var_c^\perp \cos\bigg(\f{l^2\delta S}{{2}}+\f{\phi_+^B-\phi_-^B}{2}-\bar{\varphi}\bigg) \lin
    &\approx \f{(-1)^{n+1}m\alpha^2(E-\epsilon_0)(\var_c^{\perp})^{1/2}}{2\sqrt{\pi}\hbar^2
    Z_{\parallel}^{3/2}}\cos\bigg(\f{l^2\delta S}{{2}}-\bar{\varphi}\bigg),\label{RZfirstorderE}}
with the right-hand side (of each line) evaluated at $E{=}E_n^0(B)$. The second line is an approximation for $Z_{\perp}{\rightarrow}0$, and has greater utility where $m{\ll}m_0$. $E_n^{(0)}+\delta E_n$ are sketched in the upper panel of Fig. \ref{fig:RZ}(b) as dashed lines. The validity of Eq. (\ref{RZfirstorderE}) is confirmed by a comparison between $\delta E_n$ in Eq. (\ref{RZfirstorderE}) and $\delta E_n$ extracted from numerical diagonalization in the lower panel of Fig. \ref{fig:RZ}(b).

@@

@@

We have argued in \s{sec:qtznrules} that the subleading corrections ($\lambda_{1,2}$) to the quantization rule [cf.\ \q{eq:rule}] generically lift the spin-degeneracy of Landau levels -- the level splitting is approximately $\var_c\delta_\lambda/2\pi$ with $\delta \lambda=|\lambda_1-\lambda_2|$. Conversely, Landau levels are approximately spin-degenerate if $\lambda_1{=}\lambda_2$ (mod $2\pi$) -- this degeneracy condition for the eigenvalues of $\A$ may require fine-tuning of the Peierls-Onsager Hamiltonian.

@@

In this and the subsequent section (\s{sec:rotsymmbreakdown}), we will demonstrate how rotational symmetry reduces codimension in the Landau-level problem. Before delving into this, it is instructive to consider the conceptually-simplest analog in differential geometry: if we parametrize the 2D plane by Cartesian coordinates $(x,y){\in}\R^2$, it is apparent that two parameters must be tuned to arrive at the origin. If however the origin is the center of a rotation symmetry, we may instead utilize polar coordinates and tune a single parameter (the radial coordinate) to arrive at $x{=}y{=}0$. We see that rotational symmetry reduces the codimension of the origin from two to one. Here, and in the  Landau-level problem, we define codimension with respect to a \textit{symmetrically-chosen coordinates} (if symmetry exists); such symmetric coordinates need not be well-defined at isolated high-symmetry points, just as the polar coordinates are not well-defined at the rotational center.
